Synthesizing Quotient Lenses

### 2017-10-24 papers/QuotientLensSynth/main.tex

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{canonizers-outside}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
      screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
      (if we put ``from [Nate's thesis]'' in the caption), but I don't think
      we want the full complexity of asymmetric lenses here.  We should draw
      the figure again.  (I have the OmniGraffle file for it, so this is
      easy if I know how it should look.)}\bcp{Done---what do you think?}}
  \label{fig:attheedges}
\end{figure}

@@

\begin{prooftree}
\AxiomC{$\mathcal{L}(R) \cap \mathcal{L}(R') = \varnothing$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\BinaryInfC{$squash(R, R', f)$ is well formed}
\end{prooftree}

@@

\section{QRE Lenses}
\label{QRE-lenses} 
As we have seen, QREs express a broad class of equivalence relations
directly on regular languages.  QREs are therefore a good \textit{specification}
language for quotient lenses. In this section of the paper, we introduce
\textit{QRE Lenses}. QRE lenses are quotient lenses based on the Boomerang
quotient lens combinators, but which use QREs to describe their source and view
types.

@@

\noindent One potential pitfall with QRE lenses is that the combinators which
we define on them may not preserve this normal form. For example, the
composition $q' \circ q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens which has the canonical normal form of a bijective lens with
canonizers a the edges. Consequently, in order to define quotient lenses, we
need to choose a class of bijections that has just the right properties with
respect to the quotient lens combinators. Our choice is to use the class of
\textit{bijective lenses}, a subset of the lenses expressible in Boomerang which
we identified and studied in our previous work~\cite{popl18}. Bijective lenses,
which we briefly discuss in the next subsection, possess the properties needed
to ensure that all QRE lenses have the normal form suggested in
\ref{normalform}.

\subsection{Bijective Lenses}
Define the set of \textit{bijective lenses} to be the set of bijections
between regular languages created from Boomerang lens combinators.
The syntax for the language of bijective lenses is given by
$$\ell := \mathit{id} \; (R) \sep const(s, s') \sep  swap(\ell,
\ell) \sep \ell \cdot \ell' \; |  \; (\ell \sep \ell') \sep \ell^* \;
| \; \ell \circ \ell',$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

The denotation of a lens $\ell$ is $\llbracket \ell \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell
\rrbracket$, then $\ell$ maps between $s_1$ and $s_2$.

@@

The $\mathit{const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the view data in the backward
direction. $\mathit{id}(R)$ applies the identity function to the source and view
in $\mathcal{L}(R)$ in both directions. The composition lens $\ell' \circ \ell$
applies $\ell$ followed by then $\ell'$ to the source in the forward direction,
and applies $\ell'$ followed by $\ell$ to the view in the backward direction.
The lens $\ell \cdot \ell'$ first splits the string $s$ into $s_1$ and $s_2$,
applies $\ell$ and $\ell'$ to $s_1$ and $s_2$ to get $t_1$ and $t_2$
respectively, then concatenates $t_1$ and $t_2$ and returns $t_1 \cdot t_2$ as
the final result in the forward direction. $\ell \cdot \ell'$ operates
similarly in the backward direction, but with $s, s_1$ and $s_2$ substituted
for $t, t_1$ and $t_2$. The $\mathit{swap} \; (\ell, \ell')$ lens operates
like $\ell \cdot \ell'$, except that it swaps $t_1$ and $t_2$ before
concatenating the two for a final result of $t_2 \cdot t_1$ in the forward
direction. In the backward direction, $\mathit{swap}(\ell, \ell')$ first undoes
the swap, then proceeds as expected. The $\ell \; | \; \ell'$ lens
chooses to apply $\ell$ or $\ell'$ depending on whether the source
(resp. view) data is matched by $\ell$ or $\ell'$ in the forward (resp.
backward direction. The $\ell^*$ lens splits the string $s$ into strings $s_1,
\ldots, s_n$, applies $\ell$ to each $s_i$ to get $t_i$, and then concatenates
each of the $t_i$'s for a final result of $t_1 \cdot \ldots \cdot t_n$ in the
forward direction. $\ell^*$ operates similarly in the backward direction, but
with $s, s_i$ substituted for $t, t_i$.

Each bijective lens $\ell$ has a type $\ell : R \Leftrightarrow S$ where $R$ and
$S$ are regular expressions. If $\ell : R \Leftrightarrow S$, then the source
type of $\ell$ is $\mathcal{L}(R)$ and the view type of $\ell$ is
$\mathcal{L}(S)$. Interestingly, with the bijective lens type system, a lens
$\ell : R \Leftrightarrow S$ can also be considered to be of type $\ell : R'
\Leftrightarrow S'$ provided that $R$ (resp. $S$) can be proven to be
equivalent to $R'$ (resp. $S$) from the star-semiring axioms. The typing rules
for bijective lenses are given in Figure~\ref{fig:lensrules}.

@@

\begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\AxiomC{$R \equiv R'$}
\AxiomC{$S \equiv S'$}
\TrinaryInfC{$\ell : R' \Leftrightarrow S'$}
\end{prooftree}
  \caption{Bijective Lens Typing Rules}
  \label{fig:lensrules}
\end{figure}

\subsection{Syntax of QRE Lenses}
Having given a brief overview of the class of bijective lenses, we now introduce
the class of QRE lenses, whose language is given by following grammar:
$$ q := \mathit{lift}(\ell) \sep \mathit{lquot}(q, \ell) \sep
\mathit{rquot}(\ell, q) \sep q \cdot q' \sep (q \sep q') \sep q^* \sep q \circ q',$$
where $\ell$ ranges over bijective lenses.

The QRE lens combinators are inspired by the Boomerang lens combinators of the
same name. The $\mathit{lift}(\ell)$ quotient lens enables a bijective lens
$\ell$ to be considered a quotient lens where the equivalence relation applied
to the source and view types is the equality relation, and the regular
combinators $\cdot \; * \; |$ on QRE lenses exhibit the expected behaviour,
with the caveat that the regular expressions used by these lenses are strongly
unambiguous.

The more interesting combinators are the $\mathit{lquot}$ and $\mathit{rquot}$
combinators. The $\mathit{lquot}(c, q)$ combinator takes a quotient lens $q$ and
quotients the source data using $c$, assuming that the source data forms a
complete set of representatives for the equivalence relation $\eqrel{c}$. The
$\mathit{lquot}(q, c)$ combinator does the same but on the view data. For
example, recall that in the \bibtex{} to EndNote transformation, we had the QREs
\cd{bibtex} and \cd{endnote} describing the \bibtex{} and EndNote records
respectively, and the bijective lens \cd{bib_to_end} that maps between
\cd{bibtex} and \cd{endnote}. The quotient lens which maps between
\cd{bibtex} and \cd{endnote} is given by the QRE lens

@@

The difference between QRE lenses and Boomerang quotient lenses lies in the
typing judgements. More concretely, each QRE lens $q$ has a type $q : c
\Leftrightarrow c'$ where $c, c'$ are QREs. In contrast, the approach used to
type quotient lenses in Boomerang is to classify equivalences according to
whether they are or are not the equality relation. This type system is based on
two observations: first, that most quotient lenses originate as lifted basic
lenses, and therefore have types whose equivalence relations are both equality;
and second, that equality is preserved by many of the quotient lens
combinators. Foster et al also discuss a second possible approach to typing
quotient lenses, where equivalence relations are represented by rational
functions that induce them. While this second approach is more refined than the
first, Boomerang favours the first approach since the second appears to be too
expensive to be useful in practice~\cite{quotientlenses}.

\subsection{Semantics of QRE Lenses}
The denotation $\llbracket q \rrbracket$ of a QRE lens $q:c \Leftrightarrow c'$
is a quotient lens $\llbracket q \rrbracket : W(c)/{\eqrel{c}}
\Longleftrightarrow W(c')/{\eqrel{c'}}$. The typing rules and denotation of QRE
lenses are given in Figure~\ref{fig:qlenssemantics}. The following theorem
confirms that QRE lenses exhibit the intended behaviour.

\begin{theorem}
If there is a derivation $q:c \Leftrightarrow c'$ then
$\llbracket q \rrbracket : W(c)/{\eqrel{c}} \Leftrightarrow
W(c')/{\eqrel{c'}}$ is a well-defined quotient lens.
\end{theorem}

 \begin{figure}[t]
  \centering
  \begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\UnaryInfC{$\mathit{lift}(\ell): R \Leftrightarrow id(S)$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get &=  \llbracket \ell \rrbracket, \text{ and }\\
  \llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1}
  \end{align*}

@@

  \begin{prooftree}
\AxiomC{$q_1 : c \Leftrightarrow c'$}
\AxiomC{$q_2 : c' \Leftrightarrow c''$}
\BinaryInfC{$q_2 \circ q_1: c \Leftrightarrow c''$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get &= \llbracket q_2 \rrbracket.get\circ \llbracket
  q_1 \rrbracket.get, \text{ and }\\
  \llbracket q \rrbracket.put &= \llbracket q_1 \rrbracket.put \circ \llbracket
  q_2 \rrbracket.put
  \end{align*}

@@

\subsection{Normal Forms of QRE Lenses}
Recall that our approach in defining QRE lenses is to have each QRE lens $q: c
\Leftrightarrow c'$ be such that 
\begin{align*}
\llbracket q \rrbracket.get &= \ell \circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \ell^{-1} \circ
\canonizer(c')
\end{align*}
for some bijective lens $\ell$. We now provide a proof that all QRE lenses are
of this form.\bcp{Maybe some cases of the proof can be elided in the short
  version?} 
\begin{theorem}\label{normal form}
If there is a derivation $q : c \Leftrightarrow c'$, then there exists a
bijective lens $\ell : K(c) \Leftrightarrow K(c')$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c')
\end{align*}
\end{theorem}
\begin{proof}
Assume that $q : c \Leftrightarrow c'$. We proceed by induction over the
derivation $q : c \Leftrightarrow c'$.
\begin{enumerate}
  \item
  $\mathit{lift}(\ell): R/\mathit{id}(R) \Leftrightarrow S/\mathit{id}(S)$ where
  $\ell :
  R \Leftrightarrow S$. Then
  \begin{align*}
  \llbracket \mathit{lift}(\ell) \rrbracket.get &=  \llbracket \ell \rrbracket
  = \llbracket \ell \rrbracket \circ id_{\mathcal{L}(R)} =
  \llbracket \ell \rrbracket \circ \canonizer(\mathit{id}(R)), \text{ and }\\
  \llbracket \mathit{lift}(\ell) \rrbracket.put &= \llbracket \ell
  \rrbracket^{-1} = \llbracket \ell \rrbracket^{-1} \circ id_{\mathcal{L}(S)} =
  \llbracket \ell \rrbracket^{-1} \circ \canonizer(id(S))
  \end{align*}
  \item
  $\mathit{lquot}(c, q'): c' \circ c \Leftrightarrow c''$ where $q' : c' 
  \Leftrightarrow c''$, $c$ is well formed and $K(c) = W(c')$. Then
\begin{align*}
  \llbracket q \rrbracket.get  &= \llbracket q'
  \rrbracket.get \circ \canonizer(c)\\
  \llbracket q \rrbracket.put &= \llbracket q' \rrbracket.put
  \end{align*}
  By the induction hypothesis, there exists a bijective lens $\ell :
  K(c') \Leftrightarrow K(c'')$ such that 
  \begin{align*}
\llbracket q' \rrbracket.get &= \llbracket \ell \rrbracket \circ \canonizer(c')\\
\llbracket q' \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
Canonize(c'')
\end{align*}
Consequently
\begin{align*}
  \llbracket q \rrbracket.get  &= (\llbracket \ell \rrbracket \circ
  \canonizer(c')) \circ \canonizer(c) = \llbracket \ell \rrbracket \circ
  (\canonizer(c' \circ c))\\
  \llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
  Canonize(c'')
  \end{align*}

@@

\section{Synthesizing QRE Lenses}
\label{synth}
Our goal in this section is to demonstrate how to synthesize a quotient lens
$q: c \Leftrightarrow c'$ from the QREs $c, c'$ and a set of input-output pairs
$\{(x_1, y_1), \ldots, (x_n, y_n)\}$ where the $x_i$'s are in $W(c)$ and the
$y_i$'s are in $W(c')$.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{.49\textwidth}
    \centering
    %\includegraphics{generated-graphs/times_opt}
    \caption{}
    \label{subfig:lenssize}
  \end{subfigure}
  \begin{subfigure}[b]{.49\textwidth}
    %\includegraphics{times_new.eps}
    \caption{}
    \label{subfig:examplesused}
  \end{subfigure}
  \caption{Runtimes of Optician and \Name{}.
    In (a), we run Optician and \Name{} on Optician's benchmarks.  We find that
    there is only a negligible difference incurred by Optometrist's overhead.
    In (b), we run Optometrist on its benchmark suite.  We find that it is able
    to synthesize all quotient lenses in under 20 seconds, and typically
    finished in under 5 seconds. }
  \label{fig:times}
\end{figure}

>>>>

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{canonizers-outside}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
      screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
      (if we put ``from [Nate's thesis]'' in the caption), but I don't think
      we want the full complexity of asymmetric lenses here.  We should draw
      the figure again.  (I have the OmniGraffle file for it, so this is
      easy if I know how it should look.)}\bcp{Done---what do you think?}
      \sam{Yes this is more accurate, except that we don't need a general choose
      function. With QREs the kernel language is always contained in the whole
      language so we just use the inclusion as the choose function. Also,
      perhaps it would be clearer if we said source structures and target
      structres instead of all structures?}}
  \label{fig:attheedges}
\end{figure}

@@

\begin{prooftree}
\AxiomC{$\mathcal{L}(R) \cap \mathcal{L}(R') = \varnothing$}
\AxiomC{$f : \mathcal{L}(R) \longrightarrow \mathcal{L}(R')$}
\BinaryInfC{$squash(R, R', \ell)$ is well formed}
\end{prooftree}

@@

\section{QRE Lenses}
\label{QRE-lenses} 
As we have seen, QREs express a broad class of equivalence relations
directly on regular languages.  QREs are therefore a good \textit{specification}
language for quotient lenses. In this section of the paper, we introduce
\textit{QRE Lenses}, a class of quotient lenses specified using QREs.

@@

\noindent In other words, the get function of the QRE first canonizes the source
data using $\mathit{Canonizer}(c)$ and then applies $f$ to the result, while the
put function of the QRE first canonizes the target data using
$\mathit{Canonizer}(c')$ and the applies $f^{-1}$ to the result.

One potential pitfall with QRE lenses is that the combinators which
we define on them may not preserve this normal form. For example, the
composition $q' \circ q$ of QRE lenses $q$ and $q'$ may not be equivalent to
any quotient lens which has the canonical normal form of a bijective lens with
canonizers a the edges. Consequently, in order to define quotient lenses, we
need to choose a set $S$ of bijections that has the right properties in the
sense that QRE lenses built from the bijections in $S$ are closed under the
action of QRE lens combinators. Our choice is to use the class of
\textit{bijective lenses}, which are the bijections that can be expressed in
Boomerang. In the next subsection we briefly introduce bijective lenses and
discuss the properties of bijective lenses which we will use to prove that QRE
lenses have the normal form of bijective lens with canonizers at the edges.

\subsection{Bijective Lenses}
We define the set of \textit{bijective lenses} to be the set of bijections
between regular languages created using Boomerang lens combinators.
The syntax for the language of bijective lenses is given by
$$\ell := \mathit{id} \; (R) \sep const(s, s') \sep  swap(\ell,
\ell) \sep \ell \cdot \ell' \; |  \; (\ell \sep \ell') \sep \ell^* \;
| \; \ell \circ \ell',$$ where $R$ ranges over regular expressions and $s$
ranges over character strings.

The denotation of a lens $\ell$ is $\llbracket \ell \rrbracket \subseteq
\mathit{String} \times \mathit{String}$. If $(s_1, s_2) \in \llbracket \ell
\rrbracket$, then $\ell$ maps between $s_1$ and $s_2$.

@@

The $\mathit{const}(s, t)$ lens replaces the string $s$ with $t$ in the source
data in the forward direction, and $t$ with $s$ in the target data in the backward
direction. $\mathit{id}(R)$ applies the identity function to the source and target
in $\mathcal{L}(R)$ in both directions. The composition lens $\ell' \circ \ell$
applies $\ell$ followed by then $\ell'$ to the source in the forward direction,
and applies $\ell'$ followed by $\ell$ to the target in the backward direction.
The lens $\ell \cdot \ell'$ first splits the string $s$ into $s_1$ and $s_2$,
applies $\ell$ and $\ell'$ to $s_1$ and $s_2$ to get $t_1$ and $t_2$
respectively, then concatenates $t_1$ and $t_2$ and returns $t_1 \cdot t_2$ as
the final result in the forward direction. $\ell \cdot \ell'$ operates
similarly in the backward direction, but with $s, s_1$ and $s_2$ substituted
for $t, t_1$ and $t_2$. The $\mathit{swap} \; (\ell, \ell')$ lens operates
like $\ell \cdot \ell'$, except that it swaps $t_1$ and $t_2$ before
concatenating the two for a final result of $t_2 \cdot t_1$ in the forward
direction. In the backward direction, $\mathit{swap}(\ell, \ell')$ first undoes
the swap, then proceeds as expected. The $\ell \; | \; \ell'$ lens
chooses to apply $\ell$ or $\ell'$ depending on whether the source
(resp. target) data is matched by $\ell$ or $\ell'$ in the forward (resp.
backward direction). The $\ell^*$ lens splits the string $s$ into strings $s_1,
\ldots, s_n$, applies $\ell$ to each $s_i$ to get $t_i$, and then concatenates
each of the $t_i$'s for a final result of $t_1 \cdot \ldots \cdot t_n$ in the
forward direction. $\ell^*$ operates similarly in the backward direction, but
with $s, s_i$ substituted for $t, t_i$.

Each bijective lens $\ell$ has a type $\ell : R \Leftrightarrow S$ where $R$ and
$S$ are regular expressions. If $\ell : R \Leftrightarrow S$, then the source
type of $\ell$ is $\mathcal{L}(R)$ and the target type of $\ell$ is
$\mathcal{L}(S)$. Interestingly, with the bijective lens type system, a lens
$\ell : R \Leftrightarrow S$ can also be considered to be of type $\ell : R'
\Leftrightarrow S'$ provided that $R$ (resp. $S$) can be proven to be
equivalent to $R'$ (resp. $S$) from the star-semiring axioms:

\begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\AxiomC{$R \equiv^d R'$}
\AxiomC{$S \equiv^d S'$}
\TrinaryInfC{$\ell : R' \Leftrightarrow S'$}
\end{prooftree}

The other typing rules for bijective lenses are given in
Figure~\ref{fig:lensrules}.

@@

\begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\AxiomC{$R \equiv^d R'$}
\AxiomC{$S \equiv^d S'$}
\TrinaryInfC{$\ell : R' \Leftrightarrow S'$}
\end{prooftree}
  \caption{Bijective Lens Typing Rules}
  \label{fig:lensrules}
\end{figure}

\subsection{Syntax of QRE Lenses}
Having given a brief overview of the class of bijective lenses, we now introduce
the class of QRE lenses. The language of QRE lenses is given by following
grammar:
$$ q := \mathit{lift}(\ell) \sep \mathit{lquot}(q, \ell) \sep
\mathit{rquot}(\ell, q) \sep q \cdot q' \sep (q \sep q') \sep q^* \sep q \circ q',$$
where $\ell$ ranges over bijective lenses.

The QRE lens combinators are inspired by the Boomerang lens combinators of the
same name. The $\mathit{lift}(\ell)$ quotient lens enables a bijective lens
$\ell$ to be considered a quotient lens, with the equivalence relation applied
to both the source and target formats being the equality relation. The
regular combinators $\cdot \; * \; |$ on QRE lenses exhibit the expected
behaviour, with the caveat that the regular expressions used by these lenses
are strongly unambiguous. The composition operator $\circ$ also exhibits the
expected behaviout, though the conditions that must be checked in order for the
composition to be well defined are perhaps unintuitive. The reasons why
these conditions are necesaary will become more apparent when proving normal
forms for QRE lenses.

The more interesting combinators are the $\mathit{lquot}$ and $\mathit{rquot}$
combinators. The $\mathit{lquot}(c, q)$ combinator takes a quotient lens $q$ and
quotients the source data using $c$, assuming that the source data forms a
complete set of representatives for the equivalence relation $\eqrel{c}$. The
$\mathit{lquot}(q, c)$ combinator does the same but on the view data.

For example, recall that in the \bibtex{} to EndNote transformation, we had the
QREs \cd{bibtex} and \cd{endnote} describing the \bibtex{} and EndNote records
respectively, and the bijective lens \cd{bib_to_end} that maps between
\cd{bibtex} and \cd{endnote}. The quotient lens which maps between
\cd{bibtex} and \cd{endnote} is given by the QRE lens

@@

\iffalse
The difference between QRE lenses and Boomerang quotient lenses lies in the
typing judgements. More concretely, each QRE lens $q$ has a type $q : c
\Leftrightarrow c'$ where $c, c'$ are QREs. In contrast, the approach used to
type quotient lenses in Boomerang is to classify equivalences according to
whether they are or are not the equality relation. This type system is based on
two observations: first, that most quotient lenses originate as lifted basic
lenses, and therefore have types whose equivalence relations are both equality;
and second, that equality is preserved by many of the quotient lens
combinators. Foster et al also discuss a second possible approach to typing
quotient lenses, where equivalence relations are represented by rational
functions that induce them. While this second approach is more refined than the
first, Boomerang favours the first approach since the second appears to be too
expensive to be useful in practice~\cite{quotientlenses}.
\fi

\subsection{Semantics of QRE Lenses}
Each QRE lens $q$ has a type $q : c \Leftrightarrow c'$ where $c, c'$ are QREs.
If $q : c \Leftrightarrow c'$, then the source format is described by $W(c)$ and
the canonical set of representative for the source data is described by $K(c)$.
Similarly, the target format is described by $W(c')$ and the canonical set of
representatives for the target data is described by $K(c')$. The underlying lens
of $q$ is a bijective lens $\ell : K(c) \Leftrightarrow K(c')$.

The denotation $\llbracket q \rrbracket$ of a QRE lens $q:c \Leftrightarrow c'$
is a quotient lens $\llbracket q \rrbracket : W(c)/{\eqrel{c}}
\Longleftrightarrow W(c')/{\eqrel{c'}}$. The typing rules and denotation of QRE
lenses are given in Figure~\ref{fig:qlenssemantics}.

The following theorem confirms that the semantics for QRE lenses are correct:
\begin{theorem}
If there is a derivation $q : c \Leftrightarrow c'$, then $\llbracket q
\rrbracket : W(c)/{\eqrel{c}} \Leftrightarrow W(c')/{\eqrel{c'}}$ is a
well-defined quotient lens.
\end{theorem}

 \begin{figure}[t]
  \centering
  \begin{prooftree}
\AxiomC{$\ell : R \Leftrightarrow S$}
\UnaryInfC{$\mathit{lift}(\ell): id(R) \Leftrightarrow id(S)$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get &=  \llbracket \ell \rrbracket, \text{ and }\\
  \llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1}
  \end{align*}

@@

  \begin{prooftree}
\AxiomC{$q_1 : c \Leftrightarrow c'$}
\AxiomC{$q_2 : c'' \Leftrightarrow c'''$}
\AxiomC{$\substack{\mathcal{L}(W(c'))= \mathcal{L}(W(c'')) \\ K(c') \equiv^d
K(c'')}$}
\AxiomC{$\mathit{Canonizer}(c') = \mathit{Canonizer}(c'')$}
\QuaternaryInfC{$q_2 \circ q_1: c \Leftrightarrow c'''$}
\end{prooftree}
  \begin{align*}
  \llbracket q \rrbracket.get &= \llbracket q_2 \rrbracket.get\circ \llbracket
  q_1 \rrbracket.get, \text{ and }\\
  \llbracket q \rrbracket.put &= \llbracket q_1 \rrbracket.put \circ \llbracket
  q_2 \rrbracket.put
  \end{align*}

@@

\subsection{Normal Forms of QRE Lenses}
Recall that our approach in defining QRE lenses is to have each QRE lens $q: c
\Leftrightarrow c'$ be such that 
\begin{align*}
\llbracket q \rrbracket.get &= \ell \circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \ell^{-1} \circ
\canonizer(c')
\end{align*}
for some bijective lens $\ell$. In other words, each QRE lens is the same as a
bijective lens with canonizers at the edges. The following theorem confirms that
this indeed is the case:

\begin{theorem}\label{normal form}
If there is a derivation $q : c \Leftrightarrow c'$, then there exists a
bijective lens $\ell : K(c) \Leftrightarrow K(c')$ such that
\begin{align*}
\llbracket q \rrbracket.get &= \llbracket \ell \rrbracket\circ \canonizer(c)\\
\llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
\canonizer(c')
\end{align*}
\end{theorem}
We elide a full proof of this statement, since the proof is attained via a
straightforward induction on the derivation $q : c \Leftrightarrow c$. Since
bijective lenses are closed under strongly unambiguous regular operators, the
tricky case is composition. The composition case is more involved since we need
to show that we can get rid of the canonizers in the middle:

\sam{TODO: Put a diagram here to show what you mean.}

The derivation rule and denotation for composition are as follows:
  \begin{prooftree}
\AxiomC{$q_1 : c \Leftrightarrow c'$}
\AxiomC{$q_2 : c'' \Leftrightarrow c'''$}
\AxiomC{$\substack{\mathcal{L}(W(c'))= \mathcal{L}(W(c'')) \\ K(c') \equiv^d
K(c'')}$}
\AxiomC{$\mathit{Canonizer}(c') = \mathit{Canonizer}(c'')$}
\QuaternaryInfC{$q_2 \circ q_1: c \Leftrightarrow c'''$}
\end{prooftree}

\begin{align*}
  \llbracket q \rrbracket.get &= \llbracket q_2 \rrbracket.get\circ \llbracket
  q_1 \rrbracket.get, \text{ and }\\
  \llbracket q \rrbracket.put &= \llbracket q_1 \rrbracket.put \circ \llbracket
  q_2 \rrbracket.put
  \end{align*}

This derivation rule says that the composition $q_2 \circ q_1$ is well defined
if and only if the intermediary QREs $c''$ and $c'''$ define the same
equivalence relation on the same regular language. The condition
$\mathcal{L}(W(c')) = \mathcal{L}(W(c''))$ says that the intermediary language
is the same on both sides, while the condition $K(c') \equiv^d K(c'')$ says that
the kernel regular expressions are equivalent via the star-semiring axioms.
Finally, the condition $\mathit{Canonizer}(c') = \mathit{Canonizer}(c'')$ says
that $c'$ and $c''$ define the same equivalence relation on $\mathcal{L}(W(c'))=
\mathcal{L}(W(c''))$.

  By the induction hypothesis, there exist bijective lenses
  $\ell_1 :
  K(c) \Leftrightarrow K(c')$ and $\ell_2 : K(c'') \Leftrightarrow K(c''')$ such
  that
  \begin{align*}
\llbracket q_1 \rrbracket.get &= \llbracket \ell_1 \rrbracket \circ
\canonizer(c)\\
\llbracket q_1 \rrbracket.put &= {\llbracket \ell_1 \rrbracket}^{-1} \circ
\canonizer(c')
\end{align*}
and
\begin{align*}
\llbracket q_2 \rrbracket.get &= \llbracket \ell_2 \rrbracket \circ
Canonize(c')\\
\llbracket q_2 \rrbracket.put &= {\llbracket \ell_2 \rrbracket}^{-1} \circ
Canonize(c'')
\end{align*}
Consequently,
\begin{align*}
\llbracket q_2 \rrbracket.get \circ \llbracket q_1 \rrbracket.get &=
(\llbracket \ell_2 \rrbracket \circ \canonizer(c')) \circ (\llbracket \ell_1
\rrbracket \circ \canonizer(c))\\
&= \llbracket \ell_2 \rrbracket \circ (\canonizer(c') \circ \llbracket \ell_1
\rrbracket) \circ \canonizer(c)\\
&= (\llbracket \ell_2 \rrbracket \circ \llbracket \ell_1 \rrbracket) \circ
\canonizer(c)\\
&= \llbracket \ell_2  \circ  \ell_1 \rrbracket \circ
\canonizer(c)
\end{align*} 
We are permitted to claim the third step from the second since
$\mathit{Canonizer(c')}$ is the identity function on $K(c'')$ which is
equivalent to $K(c')$ via the star semiring axioms by assumption.

A similar argument shows that 
$$\llbracket q_1 \rrbracket.put \circ \llbracket q_2 \rrbracket.put =
\llbracket \ell_2  \circ  \ell_1 \rrbracket^{-1} \circ
\canonizer(c)$$

The other cases of the proof are similar in that they follow from a
straightforward application of the induction hypothesis followed by unrolling
the equations that give the denotation for QRE lenses.
\iffalse
\begin{proof}
Assume that $q : c \Leftrightarrow c'$. We proceed by induction over the
derivation $q : c \Leftrightarrow c'$.
\begin{enumerate}
  \item
  $\mathit{lift}(\ell): R/\mathit{id}(R) \Leftrightarrow S/\mathit{id}(S)$ where
  $\ell :
  R \Leftrightarrow S$. Then
  \begin{align*}
  \llbracket \mathit{lift}(\ell) \rrbracket.get &=  \llbracket \ell \rrbracket
  = \llbracket \ell \rrbracket \circ id_{\mathcal{L}(R)} =
  \llbracket \ell \rrbracket \circ \canonizer(\mathit{id}(R)), \text{ and }\\
  \llbracket \mathit{lift}(\ell) \rrbracket.put &= \llbracket \ell
  \rrbracket^{-1} = \llbracket \ell \rrbracket^{-1} \circ id_{\mathcal{L}(S)} =
  \llbracket \ell \rrbracket^{-1} \circ \canonizer(id(S))
  \end{align*}
  \item
  $\mathit{lquot}(c, q'): c' \circ c \Leftrightarrow c''$ where $q' : c' 
  \Leftrightarrow c''$, $c$ is well formed and $K(c) = W(c')$. Then
\begin{align*}
  \llbracket q \rrbracket.get  &= \llbracket q'
  \rrbracket.get \circ \canonizer(c)\\
  \llbracket q \rrbracket.put &= \llbracket q' \rrbracket.put
  \end{align*}
  By the induction hypothesis, there exists a bijective lens $\ell :
  K(c') \Leftrightarrow K(c'')$ such that 
  \begin{align*}
\llbracket q' \rrbracket.get &= \llbracket \ell \rrbracket \circ \canonizer(c')\\
\llbracket q' \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
Canonize(c'')
\end{align*}
Consequently
\begin{align*}
  \llbracket q \rrbracket.get  &= (\llbracket \ell \rrbracket \circ
  \canonizer(c')) \circ \canonizer(c) = \llbracket \ell \rrbracket \circ
  (\canonizer(c' \circ c))\\
  \llbracket q \rrbracket.put &= \llbracket \ell \rrbracket^{-1} \circ
  Canonize(c'')
  \end{align*}

@@

\fi
\section{Synthesizing QRE Lenses}
\label{synth}
Our goal in this section is to demonstrate how to synthesize a quotient lens
$q: c \Leftrightarrow c'$ from the QREs $c, c'$ and a set of input-output pairs
$\{(x_1, y_1), \ldots, (x_n, y_n)\}$ where the $x_i$'s are in $W(c)$ and the
$y_i$'s are in $W(c')$.

@@

\begin{figure}
  \centering
  \begin{subfigure}[b]{.49\textwidth}
    \centering
    \includegraphics{generated-graphs/times_opt}
    \caption{}
    \label{subfig:lenssize}
  \end{subfigure}
  \begin{subfigure}[b]{.49\textwidth}
    \includegraphics{generated-graphs/times_new.eps}
    \caption{}
    \label{subfig:examplesused}
  \end{subfigure}
  \caption{Runtimes of Optician and \Name{}.
    In (a), we run Optician and \Name{} on Optician's benchmarks.  We find that
    there is only a negligible difference incurred by Optometrist's overhead.
    In (b), we run Optometrist on its benchmark suite.  We find that it is able
    to synthesize all quotient lenses in under 20 seconds, and typically
    finished in under 5 seconds. }
  \label{fig:times}
\end{figure}
